---
title: "Business Understanding"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r development, include=FALSE}
library(testthat)
```


```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```



## Project Name

**Taxi Drivers Net Earnings**.

## Problem Statement

The problem is that taxi drivers’ net earnings are not as high as they could be due to a lack of strategy for increasing tips.

## Project Scope

This project will be limited to Juno, Uber, Via and Lyft taxi drivers who work in New York City.

## Stakeholders

- Taxi drivers
- Taxi companies
- Customers
- NYC Taxi and Limousine Commission

## Top Process Definition

To define the elements of the process, we use a **SIPOC** diagram.

```{r}

DiagrammeR::grViz('
digraph SIPOC {
    rankdir=LR;
    node [shape=box];
    subgraph cluster_S {
        label="Suppliers";
        S1 [label="Gas Station"];
        S2 [label="Car Manufacturer"];
        S3 [label="Taxi Application"];
        S4 [label="Telecomuncation\nCompany"];
        S5 [label="Smartphone Supplier"];
        S6 [label="Maintenance\nService Providers"];
    }
    subgraph cluster_I {
        label="Inputs";
        I1 [label="Gas"];
        I2 [label="Car"];
        I3 [label="Start\nLocation"];
        I4 [label="End\nLocation"];
        I5 [label="Internet"];
        I6 [label="Smartphone"];
        I7 [label="Customer\nRequests"];
    }
    subgraph cluster_P {
        label="Process";
        P1 [label="The customer request a taxi"];
        P2 [label="The driver arrived at\nthe pick-up location"];
        P3 [label="Drivers pick the customer up"];
        P4 [label="Drivers drive to destination"];
        P5 [label="Drivers leave the customer\nat the end point"];
    }
    subgraph cluster_O {
        label="Outputs";
        O1 [label="The customer is picked up\n at start location"];
        O2 [label="The customer recives a\ntravel experience"];
        O3 [label="The Customer gets\nat end location"];
        O4 [label="Payment Received"]
    }
    subgraph cluster_C {
        label="Customers";
        C1 [label="Taxi User"];
    }
    S1 -> I1 [style=invis];
    I1 -> P1 [style=invis];
    P1 -> O1 [style=invis];
    O1 -> C1 [style=invis];
    P1 -> P2 [constraint=false];
    P2 -> P3 [constraint=false];
    P3 -> P4 [constraint=false];
    P4 -> P5 [constraint=false];
}

')
```

## Project Objective

The objective of this project is to develop a strategy to select the best paid trips possible and to increase their tips and thereby their net earnings.

## Defining Metric

Based on the current information we can say that our objective is to increase the **hourly wage** received by each taxi driver defined by the following formula.

$$
\text{Hourly Wage} = \frac{\text{Total Driver Pay} + \text{Total Tips}}{\text{Total Hours Worked}}
$$

## Data to Use

In this project, we will use a subset of the data available in the [TLC Trip Record Data](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page) from 2022 to 2023 for **High Volume For-Hire Vehicle** with the columns described in the README.md file located at the [data](https://github.com/AngelFelizR/nyc-taxi-project/tree/master/data) folder.

Based on the variables available, we can divide them in to 2 categories:

- **Available Before Arriving at the Pick-Up Location**: They will be used as the predictors to train the model.
  - hvfhs_license_num
  - dispatching_base_num
  - originating_base_num
  - request_datetime
  - PULocationID
  - DOLocationID
  - trip_miles
  - base_passenger_fare
  - shared_request_flag
  - access_a_ride_flag
  - wav_request_flag
  - wav_match_flag

- **Available After Ending the Trip**: They will be used to create the outcome variable needs to be predicted by the model.
  - on_scene_datetime
  - pickup_datetime
  - dropoff_datetime
  - trip_time
  - tolls
  - bcf
  - sales_tax
  - congestion_surcharge
  - airport_fee
  - tips
  - driver_pay
  - shared_match_flag


## Outcome to predict

To select better trips a taxi driver needs to have a clear which are the best trips to take, I mean the ones the request less time to get more money.

Based on the available data we can calculate the **profit_rate** in dollars by minute to  *minimize the time needed to make more money* and increasing in consequence the expected **hourly wage** in the long term.

$$
\text{profit rate} = \frac{\text{driver pay} + \text{tips}}{(\text{dropoff datetime}-\text{request datetime})/60}
$$


### Defining Metric's Base Line (Simulation)

Defining the baseline based on this data is a challenge as the data doesn't have any unique id to make the estimation, but we can **run a simulation** to estimate its value with a confident interval.

The simulation will be based on the following assumptions related to the taxi drivers:

- They can start to work from any zone of Manhattan, Brooklyn or Queens (the more active ones).
- They work from 8 to 12 hour every day.
- They can start to work in any month, weekday or hour.
- They just can take trips starting at the same zone they are after ending the their last trip
- The maximum waiting time before receiving a new trip request is 6 minutes.
- If a taxi driver cannot find a new trip in the first 6 minutes, he can to extend 6 minutes more and drive to the closest zone to find a new trip, but if that doesn't work in the next 6 minutes he can drive to any zone in the current Borough.

To do it, we will apply the next steps:

1. Load the environment to use.

```{r, warning=FALSE, message=FALSE}
library(here)
library(scales)
library(ggplot2)
library(data.table)
library(lubridate)
library(dplyr)
library(arrow)
library(infer)
```

2. Create a connection with the parquet files without loading them into RAM.

```{r}
NycTrips <- open_dataset(here("raw-data/trip-data"))

nrow(NycTrips) |> scales::comma() |> paste0(" rows")
```

3. Use the locations ids of all the trips we can estimate the mean distance between each pear of locations.

```{r}
LocationsMeanDistance <-
  NycTrips |>
  group_by(PULocationID, DOLocationID) |>
  summarize(trip_miles_mean = mean(trip_miles)) |>
  collect() |>
  as.data.table()

LocationsMeanDistance
```

4. Define a named vectors that can return the closest location id for all locations ids in the dataset. 

```{r}
ClosestZone <-
  
  # We are bind the same table after exchanging the id columns
  # to make sure that all ids are present in both column for any
  # pair of ids
  rbind(LocationsMeanDistance,
        LocationsMeanDistance[, .(PULocationID = DOLocationID,
                                  DOLocationID = PULocationID,
                                  trip_miles_mean)]
  
  # We need to make sure to place the lower distances at the top  
  )[order(
    # This expression help me to place distances between the same location
    # at the bottom of the list with needing to open another []
    (PULocationID == DOLocationID), 
    # Then we just need to pass the distance column
    trip_miles_mean
    ), 
    
    # This expression will select the first row of the data.table
    # for each PULocationID to avoid duplication
    .SD[1L],
    by = "PULocationID"
    
  # To end the process we just need to create named vector
  # that we will use to return the closest location for any  id
  ][, setattr(DOLocationID, "names", PULocationID)]


head(ClosestZone)
```

5. Restrict the location ids to use to the ones related to those zones.

```{r}
ValidZoneCodes <- fread(
  here("raw-data/taxi_zone_lookup.csv"),
  colClasses = c("integer",
                 "character",
                 "character",
                 "character")
)[Borough %chin% c("Manhattan", "Brooklyn", "Queens")]
```

6. Link all possible zones for each location id based on each Borough.

```{r}
BoroughZones <-
  ValidZoneCodes[, .(LocationID,
                     id_list = list(LocationID)),
                 by = "Borough"]

head(BoroughZones)
```


7. Defining when and where each simulation will start.

```{r}
# Repeating the experiment to create confident intervals
NumRepetitions <- 30L

# Defining the seed
set.seed(1558)

# Defining the starting point of each repetition
RandomStartPoint <- data.table(
  simulation_day = 1:NumRepetitions,
  seed_num = sample(1000:9000, NumRepetitions),
  PULocationID = sample(ValidZoneCodes$LocationID, NumRepetitions, replace = TRUE),
  hours_to_work = sample(8:12, NumRepetitions, replace = TRUE),
  start_time = make_datetime(
    year = sample(c(2022L, 2023L), NumRepetitions, replace = TRUE),
    month = sample(1:12, NumRepetitions, replace = TRUE),
    day = sample(1:31, NumRepetitions, replace = TRUE),
    hour = sample(0:23, NumRepetitions, replace = TRUE)
  )
)

head(RandomStartPoint)
```


8. Based on each start point simulate a day of work.

```{r function-simulate_trips}

#' Simulates a day of work
#' 
#' This function sample from a data.frame or arrow connection all the trips needed
#' to complete the a working day form taxi driver.
#'
#' @param arrow_con an arrow connection with trips to sample from
#' @param start_datetime a date with an associated time when the taxi driver start to work
#' @param start_zone a number to select to the starting zone
#' @param minutes_next_trip an integer to define the limit time needed before extending the filters
#' @param end_datetime a date with an associated time when the taxi driver stop looking for new trips
#' @param valid_end_zones a vector of number defining all possible zones to drive
#' @param closest_zone a named vector pointing the closest zone from the taxi driver is waiting for a need trip in order to start a trip
#' @param borough_zones a data.frame with the integer column LocationID with all possible zone ids and the list column id_list with contain all other zones ids related to the LocationID's borough
#' @param seed_num a single value, interpreted as an integer to make possible to reproduce random events
#'
#' @importFrom lubridate minutes as_datetime make_datetime hours year month
#' @importFrom dplyr select filter compute collect slice_sample bind_rows
#'
#' @return A data.frame
#' @export
#'
#' @examples
simulate_trips <- function(arrow_con,
                           start_datetime,
                           start_zone,
                           minutes_next_trip,
                           end_datetime,
                           valid_end_zones,
                           closest_zone,
                           borough_zones,
                           seed_num = 1234){
  
  # Confirm if will need to compute the results
  stopifnot("arrow_con is not an arrow connection" = inherits(arrow_con, "ArrowObject"))
  
  # Transforming minutes to period class
  minutes_next_trip <- lubridate::minutes(minutes_next_trip)
  
  # For the first trip
  current_time <- start_datetime
  current_zone <- start_zone

  # An empty table to bind the results
  done_trips <- data.frame(
    s_PULocationID = vector("integer"),
    s_DOLocationID = vector("integer"),
    s_request_datetime = vector("double") |> lubridate::as_datetime(),
    s_dropoff_datetime = vector("double") |> lubridate::as_datetime(),
    s_driver_pay = vector("double"),
    s_tips = vector("double")
  )

  # We need to keep adding trips until reaching the end_time 
  while(current_time <= end_datetime){

   # Filter all trips that meet the expected conditions
    # Remove any row that don't meet expectations
    trip_alternatives <-
      arrow_con |>
        dplyr::filter(year == lubridate::year(current_time),
                      month == lubridate::month(current_time),
                      PULocationID %in% current_zone,
                      DOLocationID %in% valid_end_zones,
                      request_datetime >= current_time,
                      request_datetime <= (current_time + minutes_next_trip)) |>
        dplyr::compute()

    # If we can not find any trip in the zone and time defined
    # we can move update the current time try to find trip in the closest zone
    if(length(current_zone) == 1L && nrow(trip_alternatives) == 0L){

      current_time <- current_time + minutes_next_trip
      current_zone <- c(current_zone, closest_zone[as.character(current_zone)])

    # If we couldn't find any trip to the closest zone
    # we can open the trip to any zone of the borough
    # after updating the current_time
    }else if(length(current_zone) > 1L && nrow(trip_alternatives) == 0L){

      current_time <- current_time + minutes_next_trip
      current_zone <- c(
        current_zone,
        borough_zones[borough_zones$LocationID == current_zone[1L], ][["id_list"]][[1L]]
      )

    }else{
      
      set.seed(seed_num)
      
      simulated_trip <- 
        # Once we have alternatives to select from can select one trip
        dplyr::slice_sample(trip_alternatives, n = 1L) |>
        
        # Add the prefix s_ to each column
        dplyr::select(s_PULocationID = PULocationID,
                      s_DOLocationID = DOLocationID,
                      s_request_datetime = request_datetime,
                      s_dropoff_datetime = dropoff_datetime,
                      s_driver_pay = driver_pay,
                      s_tips = tips) |>
        
        dplyr::collect()
      
      set.seed(NULL)
      
      # Updating starting point for next trip
      current_time <- simulated_trip$s_dropoff_datetime
      current_zone <- simulated_trip$s_DOLocationID

      # Adding the trip to final result
      done_trips <- dplyr::bind_rows(done_trips, simulated_trip)
      
    }

  }

  # Returning all trips
  return(done_trips)

}

```

```{r example-simulate_trips, echo=FALSE}

# Let's defining 3 zones
valid_zones <- 1:3

# Then define the closest zone to each zone
closest_zones <- c(2L, 1L, 2L)
names(closest_zones) <- valid_zones

# All the zones are from the same borough
borough_zones <- data.frame(
  LocationID = valid_zones,
  id_list = I(list(valid_zones, valid_zones, valid_zones))
)

start_datetime <- lubridate::make_datetime(2023L, 5L, 18L, 8L)
start_zone <- 1L

df <- data.frame(
  year = 2023L,
  month = 5L,
  PULocationID = c(
    
  # Let's simulate a trip in starting zone
  1L,
  
  # Then we can start in the closest zone
  2L,
  
  # Finally we can take a trip at borough level
  3L
  ),
  
  DOLocationID = c(3L, 1L, 1L),
  
  request_datetime = c(
   start_datetime + lubridate::minutes(2L),
   start_datetime + lubridate::minutes(20L + 7L),
   start_datetime + lubridate::minutes(37L + 14L)
  ),
  dropoff_datetime = c(
   start_datetime + lubridate::minutes(20L),
   start_datetime + lubridate::minutes(37L),
   lubridate::make_datetime(2023L, 5L, 18L, 9L, 1L)
  ),
  driver_pay = 10,
  tips = 2 
)

with(new.env(parent = globalenv()),
     simulate_trips(arrow::arrow_table(df),
               start_datetime = start_datetime,
               start_zone = start_zone,
               minutes_next_trip = 6L,
               end_datetime = start_datetime + lubridate::hours(1L),
               valid_end_zones = valid_zones,
               closest_zone = closest_zones,
               borough_zones = borough_zones))

```

```{r}
BaseLineSimulation <-
  RandomStartPoint[
    , simulate_trips(NycTrips,
                     start_datetime = start_time,
                     start_zone = PULocationID,
                     minutes_next_trip = 6L,
                     end_datetime = start_time + hours(hours_to_work),
                     valid_end_zones = ValidZoneCodes$LocationID,
                     closest_zone = ClosestZone,
                     borough_zones = BoroughZones,
                     seed_num = seed_num),
    by = "simulation_day"
  ][RandomStartPoint, on = "simulation_day"]
```

After simulating 30 days, we can use **bootstrap** to infer the distribution of the mean **Hourly Wage** for any day in the year by following the next steps.

1. Calculate the *Hourly Wage* for each simulated day.

```{r}
SimulationHourlyWage <-
  BaseLineSimulation[, .(`Driver Pay` = sum(s_driver_pay),
                         `Tips` = sum(s_tips),
                         `Hours Worked` = 
                           difftime(max(s_dropoff_datetime),
                                    min(s_request_datetime),
                                    units = "hours") |>
                           as.double()),
                     by = "simulation_day"]

SimulationHourlyWage[, `Hourly Wage` := (`Driver Pay` + Tips) / `Hours Worked`]

SimulationHourlyWage[order(-`Hourly Wage`)]
```

2. Then we need to resample with replacement a new 30 days hourly wage 3,000 times and calculate the mean of each resample.

```{r}
set.seed(1586)
BootstrapHourlyWage <-
  specify(SimulationHourlyWage,
          `Hourly Wage` ~ NULL) |>
  generate(reps = 3000, type = "bootstrap") |>
  calculate(stat = "mean")

BootstrapHourlyWage
```

3. Compute the 95% confident interval.

```{r}
BootstrapInterval <- 
  get_ci(BootstrapHourlyWage, 
         level = 0.95,
         type = "percentile")

BootstrapInterval
```

4. Visualize the estimated distribution.

```{r}
visualize(BootstrapHourlyWage)+
  shade_ci(endpoints = BootstrapInterval,
           color = "#2c77BF",
           fill = "#2c77BF")+
  annotate(geom = "text",
           y = 400,
           x = c(BootstrapInterval[1L][[1L]] - 0.4,
                 BootstrapInterval[2L][[1L]] + 0.4),
           label = unlist(BootstrapInterval) |> comma(accuracy = 0.01))+
  labs(title = "Mean Hourly Wage Distribution",
       subtitle = paste0("Mean: ", round(mean(SimulationHourlyWage$`Hourly Wage`), 2),
                         ", Median: ", round(median(SimulationHourlyWage$`Hourly Wage`), 2)),
       y = "Count")+
  theme_light()+
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        axis.title.x = element_blank())
```

## Business Case

As the base driver's pay increases with costs like gas, time and car's maintenance the best way to increase total earning is by increasing the amount of **tips** that drivers receive from customers.

Based on *424,389,806* trips recorded between 2022 and 2023, drivers received *$469,726,156* in tips which is only 5% of the total earnings for that year, for example if a driver improves his strategy to increase his tips to **20%** of his current earning he could be earning **`r dollar(41.69*8*5*4*0.2)` extra monthly** if he works 8 hours a day, 5 days each week and earns *$41.69* hourly.

```{r}
NycTrips |>
  summarize(number_of_trips = sum(!is.na(PULocationID)),
            trips_with_tips = sum(tips > 0, na.rm = TRUE),
            driver_net_earning = sum(driver_pay + tips, na.rm = TRUE),
            tips = sum(tips, na.rm = TRUE)) |>
  collect() |>
  as.data.table() |>
  (\(dt)  dt[, .(number_of_trips = comma(number_of_trips),
                 trips_with_tips = comma(trips_with_tips),
                 trips_with_tips_pct = percent(trips_with_tips / number_of_trips),
                 driver_net_earning = dollar(driver_net_earning),
                 tips = dollar(tips),
                 tips_pct = percent(tips/driver_net_earning))
          ][, melt(.SD, 
                   measure.vars = names(.SD),
                   variable.name = "Summary Variable",
                   value.name = "Total",
                   variable.factor = FALSE)]
   )()

```

## Project's side benefits

It's also important to consider that Taxi companies and customers can both benefit from drivers earning more tips in several ways:

1. **Taxi Companies**: 
    - **Employee Satisfaction**: Higher tips can lead to increased job satisfaction among drivers, which can improve their performance and reduce turnover rates.
    - **Company Reputation**: If drivers are earning more tips, it could indicate that they are providing excellent service, which can enhance the company's reputation.
    - **Customer Retention**: Satisfied drivers are more likely to provide better customer service, which can lead to higher customer retention rates.

2. **Customers**:
    - **Better Service**: Drivers who earn more tips are often those who provide better service. This could mean cleaner vehicles, more courteous behavior, and a more enjoyable ride overall.
    - **Driver Availability**: If the tip earnings are high, it could attract more drivers to work, potentially reducing wait times for customers.
    - **Safety**: Drivers who are not worried about their earnings might be less likely to engage in risky behaviors (like speeding or working overly long shifts) to earn more.

## Deliverables

A **Shiny app** which assists the drivers in focusing their attention on the better trips.

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_business_understending.Rmd",
               vignette_name = "02 - Business Understanding",
               overwrite = "yes",
               vignettes = FALSE)
```
