---
title: "02 - Business Understanding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02-business-understanding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Project Name

**Increasing NYC Taxi Drivers Earnings**.


## Problem Statement

The problem is that taxi drivers’ net earnings are not as high as they could be due to a lack of strategy for increasing tips.


## Project Scope

This project will be limited to Juno, Uber, Via and Lyft taxi drivers who work in New York City.


## Stakeholders

- Taxi drivers
- Taxi companies
- Customers
- NYC Taxi and Limousine Commission


## Top Process Definition

To define the elements of the process, we use a **SIPOC** diagram.


```{r parsermd-chunk-1, echo = FALSE}
DiagrammeR::grViz('
digraph SIPOC {
    rankdir=LR;
    node [shape=box];
    subgraph cluster_S {
        label="Suppliers";
        S1 [label="Gas Station"];
        S2 [label="Car Manufacturer"];
        S3 [label="Taxi Application"];
        S4 [label="Telecomuncation\nCompany"];
        S5 [label="Smartphone Supplier"];
        S6 [label="Maintenance\nService Providers"];
    }
    subgraph cluster_I {
        label="Inputs";
        I1 [label="Gas"];
        I2 [label="Car"];
        I3 [label="Start\nLocation"];
        I4 [label="End\nLocation"];
        I5 [label="Internet"];
        I6 [label="Smartphone"];
        I7 [label="Customer\nRequests"];
    }
    subgraph cluster_P {
        label="Process";
        P1 [label="The customer request a taxi"];
        P2 [label="The driver arrived at\nthe pick-up location"];
        P3 [label="Drivers pick the customer up"];
        P4 [label="Drivers drive to destination"];
        P5 [label="Drivers leave the customer\nat the end point"];
    }
    subgraph cluster_O {
        label="Outputs";
        O1 [label="The customer is picked up\n at start location"];
        O2 [label="The customer recives a\ntravel experience"];
        O3 [label="The Customer gets\nat end location"];
        O4 [label="Payment Received"]
    }
    subgraph cluster_C {
        label="Customers";
        C1 [label="Taxi User"];
    }
    S1 -> I1 [style=invis];
    I1 -> P1 [style=invis];
    P1 -> O1 [style=invis];
    O1 -> C1 [style=invis];
    P1 -> P2 [constraint=false];
    P2 -> P3 [constraint=false];
    P3 -> P4 [constraint=false];
    P4 -> P5 [constraint=false];
}

')
```

## Data to Use

In this project, we will use a subset of the data available in the [TLC Trip Record Data](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page) from 2022 to 2023 for **High Volume For-Hire Vehicle** with the columns described below as describes the [source data dictonary](https://www.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_hvfhs.pdf).

|**Field Name**|**Description**|
|:-------------|:--------------|
|**hvfhs_license_num**|The TLC license number of the HVFHS base or business. As of September 2019, the HVFHS licensees are the following: <br> - HV0002: Juno <br> - HV0003: Uber <br> - HV0004: Via <br> - HV0005: Lyft|
|**dispatching_base_num**|The TLC Base License Number of the base that dispatched the trip|
|**originating_base_num**|Base number of the base that received the original trip request|
|**request_datetime**|Date/time when passenger requested to be picked up|
|**on_scene_datetime**|Date/time when driver arrived at the pick-up location (Accessible Vehicles-only)|
|**pickup_datetime**|The date and time of the trip pick-up|
|**dropoff_datetime**|The date and time of the trip drop-off|
|**PULocationID**|TLC Taxi Zone in which the trip began|
|**DOLocationID**|TLC Taxi Zone in which the trip ended|
|**trip_miles**|Total miles for passenger trip|
|**trip_time**|Total time in seconds for passenger trip|
|**base_passenger_fare**|Base passenger fare before tolls, tips, taxes, and fees|
|**tolls**|Total amount of all tolls paid in trip|
|**bcf**|Total amount collected in trip for Black Car Fund|
|**sales_tax**|Total amount collected in trip for NYS sales tax|
|**congestion_surcharge**|Total amount collected in trip for NYS congestion surcharge|
|**airport_fee**|$2.50 for both drop off and pick up at LaGuardia, Newark, and John F. Kennedy airports|
|**tips**|Total amount of tips received from passenger|
|**driver_pay**|Total driver pay (not including tolls or tips and net of commission, surcharges, or taxes)|
|**shared_request_flag**|Did the passenger agree to a shared/pooled ride, regardless of whether they were matched? (Y/N)|
|**shared_match_flag**|Did the passenger share the vehicle with another passenger who booked separately at any point during the trip? (Y/N)|
|**access_a_ride_flag**|Was the trip administered on behalf of the Metropolitan Transportation Authority (MTA)? (Y/N)|
|**wav_request_flag**|Did the passenger request a wheelchair-accessible vehicle (WAV)? (Y/N)|
|**wav_match_flag**|Did the trip occur in a wheelchair-accessible vehicle (WAV)? (Y/N)|

Based on the variables available, we can divide them in to 2 categories.


### Available Before Arriving at the Pick-Up Location

They will be used as the predictors to train the model.

- ***request_datetime***
- ***PULocationID***
- ***DOLocationID***
- *hvfhs_license_num*
- *dispatching_base_num*
- *originating_base_num*
- *trip_miles*
- *base_passenger_fare*
- *shared_request_flag*
- *access_a_ride_flag*
- *wav_request_flag*
- *wav_match_flag*



### Available After Ending the Trip

They will be used to create the outcome variable needs to be predicted by the model.

- ***dropoff_datetime***
- ***driver_pay***
- ***tips***
- *on_scene_datetime*
- *pickup_datetime*
- *trip_time*
- *tolls*
- *bcf*
- *sales_tax*
- *congestion_surcharge*
- *airport_fee*
- *shared_match_flag*



## Project Objective

The objective of this project is to develop a strategy to select the best paid trips possible to get an consistent increase in NYC taxi drivers' income. 


## Defining Business Metric

Based on the current information, we can say that our objective is to increase the **Daily Hourly Wage** received by each taxi driver defined by the following formula:

$$
\text{Daily Hourly Wage} = \frac{\text{Total Earnings}}{\text{Total Hours Worked}}
$$




## Defining Outcome to Predict

Once we have a clear business problem to solve, it's important to also make the translation into a *data science* problem that we can solve using the available data.

As we want to increase earning and reduce time, we can use calculate and predict the `profit_rate` of each trip, to understand **what is behind the best pay trips**.

$$
\text{profit rate} = \frac{\text{driver pay} + \text{tips}}{\text{dropoff datetime}-\text{request datetime}}
$$



## Defining Metric's Base Line 

Defining the baseline based on this data is a challenge as the data doesn't have any *unique id* to make the estimation, but we can **run a simulation** to estimate its value with a confident interval.

The simulation will be based on the following assumptions related to the taxi drivers:

- They can start to work from any zone of Manhattan, Brooklyn or Queens (the more active ones).
- They work from 8 to 12 hour every day.
- They can start to work in any month, weekday or hour.
- They just can take trips starting at the same zone they are after ending the their last trip
- The maximum waiting time before receiving a new trip request is 6 minutes.
- If a taxi driver cannot find a new trip in the first 6 minutes, he can to extend 6 minutes more and drive to the closest zone to find a new trip, but if that doesn't work in the next 6 minutes he can drive to any zone in the current Borough.



### Runing trips simulation

1. Load the environment to use.


```{r parsermd-chunk-2, warning = FALSE, message = FALSE}
library(project.nyc.taxi)
library(here)
library(scales)
library(ggplot2)
library(data.table)
library(lubridate)
library(dplyr)
library(arrow)
library(infer)

options(datatable.print.nrows = 15)
```

2. Create a connection with the parquet files without loading them into RAM.


```{r parsermd-chunk-3}
NycTrips <- open_dataset(here("raw-data/trip-data"))

nrow(NycTrips) |> scales::comma() |> paste0(" rows")
```

3. Use the locations ids of all the trips we can estimate the mean distance between each pear of locations.


```{r parsermd-chunk-4}
LocationsMeanDistance <-
  NycTrips |>
  group_by(PULocationID, DOLocationID) |>
  summarize(trip_miles_mean = mean(trip_miles)) |>
  collect() |>
  as.data.table()

LocationsMeanDistance
```

4. Define a named vectors that can return the closest location id for all locations ids in the dataset. 


```{r parsermd-chunk-5}
ClosestZone <-
  
  # We are bind the same table after exchanging the id columns
  # to make sure that all ids are present in both column for any
  # pair of ids
  rbind(LocationsMeanDistance,
        LocationsMeanDistance[, .(PULocationID = DOLocationID,
                                  DOLocationID = PULocationID,
                                  trip_miles_mean)]
  
  # We need to make sure to place the lower distances at the top  
  )[order(
    # This expression help me to place distances between the same location
    # at the bottom of the list with needing to open another []
    (PULocationID == DOLocationID), 
    # Then we just need to pass the distance column
    trip_miles_mean
    ), 
    
    # This expression will select the first row of the data.table
    # for each PULocationID to avoid duplication
    .SD[1L],
    by = "PULocationID"
    
  # To end the process we just need to create named vector
  # that we will use to return the closest location for any  id
  ][, setattr(DOLocationID, "names", PULocationID)]


head(ClosestZone)
```

5. Restrict the location ids to use to the ones related to those zones.


```{r parsermd-chunk-6}
ValidZoneCodes <- fread(
  here("raw-data/taxi_zone_lookup.csv"),
  colClasses = c("integer",
                 "character",
                 "character",
                 "character")
)[Borough %chin% c("Manhattan", "Brooklyn", "Queens")]
```

6. Link all possible zones for each location id based on each Borough.


```{r parsermd-chunk-7}
BoroughZones <-
  ValidZoneCodes[, .(LocationID,
                     id_list = list(LocationID)),
                 by = "Borough"]

head(BoroughZones)
```

7. Defining when and where each simulation will start.


```{r parsermd-chunk-8}
# Repeating the experiment to create confident intervals
NumRepetitions <- 30L

# Defining the seed
set.seed(1558)

# Defining the starting point of each repetition
RandomStartPoint <- data.table(
  simulation_day = 1:NumRepetitions,
  PULocationID = sample(ValidZoneCodes$LocationID, NumRepetitions, replace = TRUE),
  hours_to_work = sample(8:12, NumRepetitions, replace = TRUE),
  start_time = make_datetime(
    year = sample(c(2022L, 2023L), NumRepetitions, replace = TRUE),
    month = sample(1:12, NumRepetitions, replace = TRUE),
    day = sample(1:31, NumRepetitions, replace = TRUE),
    hour = sample(0:23, NumRepetitions, replace = TRUE)
  )
)

head(RandomStartPoint)
```

8. Based on each start point simulate a day of work.

```{r parsermd-chunk-9, eval = FALSE}
BaseLineSimulation <-
  RandomStartPoint[
    , simulate_trips(NycTrips,
                     start_datetime = start_time,
                     start_zone = PULocationID,
                     minutes_next_trip = 6L,
                     end_datetime = start_time + hours(hours_to_work),
                     valid_end_zones = ValidZoneCodes$LocationID,
                     closest_zone = ClosestZone,
                     borough_zones = BoroughZones),
    by = "simulation_day"
  ][RandomStartPoint, on = "simulation_day"]
```

### Infering Daily Hourly Wage

After simulating 30 days, we can use **bootstrap** to infer the distribution of the mean **Daily Hourly Wage** for any day in the year by following the next steps.

1. Calculate the *Daily Hourly Wage* for each simulated day.


```{r parsermd-chunk-10}
SimulationHourlyWage <-
  BaseLineSimulation[, .(`Driver Pay` = sum(s_driver_pay),
                         `Tips` = sum(s_tips),
                         `Hours Worked` = 
                           difftime(max(s_dropoff_datetime),
                                    min(s_request_datetime),
                                    units = "hours") |>
                           as.double()),
                     by = "simulation_day"]

SimulationHourlyWage[, `Daily Hourly Wage` := (`Driver Pay` + Tips) / `Hours Worked`]

SimulationHourlyWage[order(-`Daily Hourly Wage`)]
```

2. Then we need to resample with replacement a new 30 days hourly wage 3,000 times and calculate the mean of each resample.


```{r parsermd-chunk-11}
set.seed(1586)
BootstrapHourlyWage <-
  specify(SimulationHourlyWage,
          `Daily Hourly Wage` ~ NULL) |>
  generate(reps = 3000, type = "bootstrap") |>
  calculate(stat = "mean")

BootstrapHourlyWage
```

3. Compute the 95% confident interval.


```{r parsermd-chunk-12}
BootstrapInterval <- 
  get_ci(BootstrapHourlyWage, 
         level = 0.95,
         type = "percentile")

BootstrapInterval
```

4. Visualize the estimated distribution.


```{r parsermd-chunk-13}
visualize(BootstrapHourlyWage)+
  shade_ci(endpoints = BootstrapInterval,
           color = "#2c77BF",
           fill = "#2c77BF")+
  annotate(geom = "text",
           y = 400,
           x = c(BootstrapInterval[1L][[1L]] - 0.4,
                 BootstrapInterval[2L][[1L]] + 0.4),
           label = unlist(BootstrapInterval) |> comma(accuracy = 0.01))+
  labs(title = "Mean Hourly Wage Distribution",
       subtitle = paste0("Mean: ", round(mean(SimulationHourlyWage$`Daily Hourly Wage`), 2),
                         ", Median: ", round(median(SimulationHourlyWage$`Daily Hourly Wage`), 2)),
       y = "Count")+
  theme_light()+
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        axis.title.x = element_blank())
```

## Business Case

As the base driver's pay increases with costs like gas, time and car's maintenance the best way to increase total earning is by increasing the amount of **tips** that drivers receive from customers.

Based on *424,389,806* trips recorded between 2022 and 2023, drivers received *$469,726,156* in tips which is only 5% of the total earnings for that year, for example if a driver improves his strategy to increase his tips to **20%** of his current earning he could be earning **`r dollar(41.69*8*5*4*0.2)` extra monthly** if he works 8 hours a day, 5 days each week and earns *$41.69* hourly.


```{r parsermd-chunk-14}
NycTrips |>
  summarize(number_of_trips = sum(!is.na(PULocationID)),
            trips_with_tips = sum(tips > 0, na.rm = TRUE),
            driver_net_earning = sum(driver_pay + tips, na.rm = TRUE),
            tips = sum(tips, na.rm = TRUE)) |>
  collect() |>
  as.data.table() |>
  (\(dt)  dt[, .(number_of_trips = comma(number_of_trips),
                 trips_with_tips = comma(trips_with_tips),
                 trips_with_tips_pct = percent(trips_with_tips / number_of_trips),
                 driver_net_earning = dollar(driver_net_earning),
                 tips = dollar(tips),
                 tips_pct = percent(tips/driver_net_earning))
          ][, melt(.SD, 
                   measure.vars = names(.SD),
                   variable.name = "Summary Variable",
                   value.name = "Total",
                   variable.factor = FALSE)]
   )()

```

## Project's Side Benefits

It's also important to consider that Taxi companies and customers can both benefit from drivers earning more tips in several ways:

1. **Taxi Companies**: 
    - **Employee Satisfaction**: Higher tips can lead to increased job satisfaction among drivers, which can improve their performance and reduce turnover rates.
    - **Company Reputation**: If drivers are earning more tips, it could indicate that they are providing excellent service, which can enhance the company's reputation.
    - **Customer Retention**: Satisfied drivers are more likely to provide better customer service, which can lead to higher customer retention rates.

2. **Customers**:
    - **Better Service**: Drivers who earn more tips are often those who provide better service. This could mean cleaner vehicles, more courteous behavior, and a more enjoyable ride overall.
    - **Driver Availability**: If the tip earnings are high, it could attract more drivers to work, potentially reducing wait times for customers.
    - **Safety**: Drivers who are not worried about their earnings might be less likely to engage in risky behaviors (like speeding or working overly long shifts) to earn more.


## Deliverables

A **Shiny app** which assists the drivers in focusing their attention on the better trips.


