---
title: "Business Understanding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Business Understanding}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
params:
  UpdateSimulation: FALSE
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library('project.nyc.taxi')
```

## Project Name

**Increasing NYC Taxi Drivers Earnings**.


## Problem Statement

The problem is that taxi drivers’ net earnings are not as high as they could be due to a lack of strategy for increasing tips.


## Project Scope

This project will be limited to Juno, Uber, Via and Lyft taxi drivers who work in New York City.


## Stakeholders

- Taxi drivers
- Taxi companies
- Customers
- NYC Taxi and Limousine Commission


## Top Process Definition

To define the elements of the process, we use a **SIPOC** diagram.


```{r parsermd-chunk-1, echo = FALSE}
DiagrammeR::grViz('
digraph SIPOC {
    rankdir=LR;
    node [shape=box];
    subgraph cluster_S {
        label="Suppliers";
        S1 [label="Gas Station"];
        S2 [label="Car Manufacturer"];
        S3 [label="Taxi Application"];
        S4 [label="Telecomuncation\nCompany"];
        S5 [label="Smartphone Supplier"];
        S6 [label="Maintenance\nService Providers"];
    }
    subgraph cluster_I {
        label="Inputs";
        I1 [label="Gas"];
        I2 [label="Car"];
        I3 [label="Start\nLocation"];
        I4 [label="End\nLocation"];
        I5 [label="Internet"];
        I6 [label="Smartphone"];
        I7 [label="Customer\nRequests"];
    }
    subgraph cluster_P {
        label="Process";
        P1 [label="The customer request a taxi"];
        P2 [label="The driver arrived at\nthe pick-up location"];
        P3 [label="Drivers pick the customer up"];
        P4 [label="Drivers drive to destination"];
        P5 [label="Drivers leave the customer\nat the end point"];
    }
    subgraph cluster_O {
        label="Outputs";
        O1 [label="The customer is picked up\n at start location"];
        O2 [label="The customer recives a\ntravel experience"];
        O3 [label="The Customer gets\nat end location"];
        O4 [label="Payment Received"]
    }
    subgraph cluster_C {
        label="Customers";
        C1 [label="Taxi User"];
    }
    S1 -> I1 [style=invis];
    I1 -> P1 [style=invis];
    P1 -> O1 [style=invis];
    O1 -> C1 [style=invis];
    P1 -> P2 [constraint=false];
    P2 -> P3 [constraint=false];
    P3 -> P4 [constraint=false];
    P4 -> P5 [constraint=false];
}

')
```

## Data to Use

In this project, we will use a subset of the data available in the [TLC Trip Record Data](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page) from 2022 to 2023 for **High Volume For-Hire Vehicle** with the columns described below as describes the [source data dictonary](https://www.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_hvfhs.pdf).

|**Field Name**|**Description**|
|:-------------|:--------------|
|**hvfhs_license_num**|The TLC license number of the HVFHS base or business. As of September 2019, the HVFHS licensees are the following: <br> - HV0002: Juno <br> - HV0003: Uber <br> - HV0004: Via <br> - HV0005: Lyft|
|**dispatching_base_num**|The TLC Base License Number of the base that dispatched the trip|
|**originating_base_num**|Base number of the base that received the original trip request|
|**request_datetime**|Date/time when passenger requested to be picked up|
|**on_scene_datetime**|Date/time when driver arrived at the pick-up location (Accessible Vehicles-only)|
|**pickup_datetime**|The date and time of the trip pick-up|
|**dropoff_datetime**|The date and time of the trip drop-off|
|**PULocationID**|TLC Taxi Zone in which the trip began|
|**DOLocationID**|TLC Taxi Zone in which the trip ended|
|**trip_miles**|Total miles for passenger trip|
|**trip_time**|Total time in seconds for passenger trip|
|**base_passenger_fare**|Base passenger fare before tolls, tips, taxes, and fees|
|**tolls**|Total amount of all tolls paid in trip|
|**bcf**|Total amount collected in trip for Black Car Fund|
|**sales_tax**|Total amount collected in trip for NYS sales tax|
|**congestion_surcharge**|Total amount collected in trip for NYS congestion surcharge|
|**airport_fee**|$2.50 for both drop off and pick up at LaGuardia, Newark, and John F. Kennedy airports|
|**tips**|Total amount of tips received from passenger|
|**driver_pay**|Total driver pay (not including tolls or tips and net of commission, surcharges, or taxes)|
|**shared_request_flag**|Did the passenger agree to a shared/pooled ride, regardless of whether they were matched? (Y/N)|
|**shared_match_flag**|Did the passenger share the vehicle with another passenger who booked separately at any point during the trip? (Y/N)|
|**access_a_ride_flag**|Was the trip administered on behalf of the Metropolitan Transportation Authority (MTA)? (Y/N)|
|**wav_request_flag**|Did the passenger request a wheelchair-accessible vehicle (WAV)? (Y/N)|
|**wav_match_flag**|Did the trip occur in a wheelchair-accessible vehicle (WAV)? (Y/N)|

Based on the variables available, we can divide them in to 2 categories.


### Available Before Arriving at the Pick-Up Location

They will be used as the predictors to train the model and can be divided into 2 groups:

1. Fixed during the experiment as they can not be changed from trip to trip.
  - *hvfhs_license_num*
  - *wav_match_flag*

2. Valid to be changed from trip to trip.
- ***request_datetime***
- ***PULocationID***
- ***DOLocationID***
- *trip_miles*
- *shared_request_flag*
- *access_a_ride_flag*
- *wav_request_flag*
- *dispatching_base_num*
- *originating_base_num*


### Available After Ending the Trip

They will be used to create the outcome variable needs to be predicted by the model.

- ***dropoff_datetime***
- ***driver_pay***
- ***tips***
- *base_passenger_fare*
- *on_scene_datetime*
- *pickup_datetime*
- *trip_time*
- *tolls*
- *bcf*
- *sales_tax*
- *congestion_surcharge*
- *airport_fee*
- *shared_match_flag*


## Project Objective

The objective of this project is to develop a strategy to select the best paid trips possible to get an consistent increase in NYC taxi drivers' income. 


## Defining Business Metric

Based on the current information, we can say that our objective is to increase the **Daily Hourly Wage** received by each taxi driver defined by the following formula:

$$
\text{Daily Hourly Wage} = \frac{\text{Total Earnings}}{\text{Total Hours Worked}}
$$

## Defining Metric's Base Line 

Defining the baseline based on this data is a challenge as the data doesn't have any *unique id* to make the estimation, but we can **run a simulation** to estimate its value with a confident interval.

The simulation will be based on the following **assumptions** related to the taxi drivers:

1. They can start to work:

- From any zone of Manhattan, Brooklyn or Queens (the more active ones)
- From any month, weekday or hour.
    
2. The TLC license number (taxi company) needs to keep constant for all trips in work day.

3. Only wheelchair-accessible vehicles can accept trips with that request.

4. As we cannot estimate the waiting time due high offer of taxis, so we will assume that **they will be able to accept the first requested trip**.

5. The taxis will find trips based on their time waiting and will take the first trip in their valid radius:

- 0-1 Minute: Search within a 1-mile radius.
- 1-3 Minutes: Expand search to a 3-mile radius if no trip is found.
- 3-5 Minutes: Expand search to a 5-mile radius if still no trip is found.
- >5 Minutes: Expand search to a 7-mile radius if no trip is found yet.

6. They have a 30 minutes break after 4 hours working once ending the current trip.

7. They will take their last trip after working 8 hours.

### Running trips simulation

1. Loading the functions to use.

```{r warning = FALSE, message = FALSE}
library(here)
library(scales)
library(ggplot2)
library(data.table)
library(lubridate)
library(infer)
library(DBI)
library(duckdb)
library(glue)

options(datatable.print.nrows = 15)
```

2. Importing the definition of each code zone.

```{r}
ZoneCodesRef <- fread(
  here("raw-data/taxi_zone_lookup.csv"),
  colClasses = c("integer",
                 "character",
                 "character",
                 "character"),
  data.table = TRUE,
  key = "Borough"
)[, Address := paste(Zone,
                     Borough,
                     "New York",
                     "United States",
                     sep = ", ")
][, tidygeocoder::geocode(.SD,
                          address = Address,
                          method = 'arcgis')]

dbWriteTable(con, "ZoneCodesRef", ZoneCodesRef)

setDT(ZoneCodesArcgis)

glimpse(ZoneCodesArcgis)
```

3. Creating and saving a temporal table listing all possible combinations of locations in **Manhattan**, **Brooklyn** and **Queens** (they have more trips that other borough in NYC) to be used as filter.

```{r}
ValidZoneCodes <-
  ZoneCodes[c("Manhattan", "Brooklyn", "Queens"), 
            on = "Borough",
            CJ(PULocationID = LocationID,
               DOLocationID = LocationID)]

dbWriteTable(con, "ValidZoneCodes", ValidZoneCodes)
```

4. Creating a connection with DuckDB.

```{r}
con <- dbConnect(duckdb(), dbdir = here("my-db.duckdb"))
```

5. Selecting at random the first trip of each simulation. It's important to know that even after setting the seed 3518 **the sample is not reproducible**, so I saved the data as part of the package of this project, so you have a easy access by running `proyect.nyc.taxi::SimulationStartDay`, why creating the `trip_id` column was so important.

```{r eval=FALSE}
SimulationStartDayQuery <- "
SELECT t1.*
FROM NycTrips t1
INNER JOIN ValidZoneCodes t2
  ON t1.PULocationID = t2.PULocationID AND
     t1.DOLocationID = t2.DOLocationID
WHERE t1.year = 2023
USING SAMPLE reservoir(60 ROWS) REPEATABLE (3518);
"

SimulationStartDay <- dbGetQuery(con, SimulationStartDayQuery)

setDT(SimulationStartDay)

pillar::glimpse(SimulationStartDay)
```

```{r include=FALSE, eval=FALSE}
usethis::use_data(SimulationStartDay,
                  overwrite = TRUE,
                  compress = "xz",
                  version = 3)
```

```{r include=FALSE}
pillar::glimpse(SimulationStartDay)
```

We can also confirm that the sample satisfy the initial restrictions:

- All trips are from 2023.

```{r}
SimulationStartDay[, .N, year]
```

- The trips begin on the expected boroughs.

```{r}
ZoneCodes[SimulationStartDay, 
          on = c("LocationID" = "PULocationID"),
          .N,
          by = "Borough"]
```

- The trips end on the expected boroughs.

```{r}
ZoneCodes[SimulationStartDay, 
          on = c("LocationID" = "DOLocationID"),
          .N,
          by = "Borough"]
```

Now we can conclude that the initial data **satisfy the assumption 1**.

6. Calculating the mean distance present from one location to other if it has **less than 7 miles**.

```{r}
MeanDistanceQuery <- "
CREATE TABLE PointMeanDistance AS

-- Selecting all avaiable from trips that don't start and end at same point
WITH ListOfPoints AS (
  SELECT 
    t1.PULocationID,
    t1.DOLocationID,
    AVG(t1.trip_miles) AS trip_miles_mean
  FROM 
    NycTrips t1
  INNER JOIN
    ValidZoneCodes t2
    ON t1.PULocationID = t2.PULocationID AND
       t1.DOLocationID = t2.DOLocationID
  WHERE
    t1.PULocationID <> t1.DOLocationID AND
    t1.year = 2023
  GROUP BY 
    t1.PULocationID, 
    t1.DOLocationID
  HAVING 
    AVG(t1.trip_miles) <= 7
),

-- Defining all available distances
ListOfPointsComplete AS (
  SELECT
    PULocationID,
    DOLocationID,
    trip_miles_mean
  FROM ListOfPoints
  UNION ALL
  SELECT
    DOLocationID AS PULocationID,
    PULocationID AS DOLocationID,
    trip_miles_mean
  FROM ListOfPoints
),
NumeredRows AS (
  SELECT
    PULocationID,
    DOLocationID,
    trip_miles_mean,
    row_number() OVER (PARTITION BY PULocationID, DOLocationID) AS n_row
  FROM ListOfPointsComplete
)

-- Selecting the first combination of distances
SELECT 
  PULocationID,
  DOLocationID,
  trip_miles_mean
FROM NumeredRows
WHERE n_row = 1
ORDER BY PULocationID, trip_miles_mean;
"

dbExecute(con, MeanDistanceQuery) |> comma()

dbListTables(con)
```


### Infering Daily Hourly Wage

After simulating 30 days, we can use **bootstrap** to infer the distribution of the mean **Daily Hourly Wage** for any day in the year by following the next steps.

1. Calculate the *Daily Hourly Wage* for each simulated day.


```{r parsermd-chunk-10}
SimulationHourlyWage <-
  BaseLineSimulation[, .(`Driver Pay` = sum(s_driver_pay),
                         `Tips` = sum(s_tips),
                         `Hours Worked` = 
                           difftime(max(s_dropoff_datetime),
                                    min(s_request_datetime),
                                    units = "hours") |>
                           as.double()),
                     by = "simulation_day"]

SimulationHourlyWage[, `Daily Hourly Wage` := (`Driver Pay` + Tips) / `Hours Worked`]

SimulationHourlyWage[order(-`Daily Hourly Wage`)]
```

2. Then we need to resample with replacement a new 30 days hourly wage 3,000 times and calculate the mean of each resample.


```{r parsermd-chunk-11}
set.seed(1586)
BootstrapHourlyWage <-
  specify(SimulationHourlyWage,
          `Daily Hourly Wage` ~ NULL) |>
  generate(reps = 3000, type = "bootstrap") |>
  calculate(stat = "mean")

BootstrapHourlyWage
```

3. Compute the 95% confident interval.


```{r parsermd-chunk-12}
BootstrapInterval <- 
  get_ci(BootstrapHourlyWage, 
         level = 0.95,
         type = "percentile")

BootstrapInterval
```

4. Visualize the estimated distribution.


```{r parsermd-chunk-13}
visualize(BootstrapHourlyWage)+
  shade_ci(endpoints = BootstrapInterval,
           color = "#2c77BF",
           fill = "#2c77BF")+
  annotate(geom = "text",
           y = 400,
           x = c(BootstrapInterval[1L][[1L]] - 0.4,
                 BootstrapInterval[2L][[1L]] + 0.4),
           label = unlist(BootstrapInterval) |> comma(accuracy = 0.01))+
  labs(title = "Mean Hourly Wage Distribution",
       subtitle = paste0("Mean: ", round(mean(SimulationHourlyWage$`Daily Hourly Wage`), 2),
                         ", Median: ", round(median(SimulationHourlyWage$`Daily Hourly Wage`), 2)),
       y = "Count")+
  theme_light()+
  theme(panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold"),
        axis.title.x = element_blank())
```

## Business Case

As the base driver's pay increases with costs like gas, time and car's maintenance the best way to increase total earning is by increasing the amount of **tips** that drivers receive from customers.

Based on *424,389,806* trips recorded between 2022 and 2023, drivers received *$469,726,156* in tips which is only 5% of the total earnings for that year, for example if a driver improves his strategy to increase his tips to **20%** of his current earning he could be earning **`r dollar(41.69*8*5*4*0.2)` extra monthly** if he works 8 hours a day, 5 days each week and earns *$41.69* hourly.


```{r parsermd-chunk-14}
NycTrips |>
  summarize(number_of_trips = sum(!is.na(PULocationID)),
            trips_with_tips = sum(tips > 0, na.rm = TRUE),
            driver_net_earning = sum(driver_pay + tips, na.rm = TRUE),
            tips = sum(tips, na.rm = TRUE)) |>
  collect() |>
  as.data.table() |>
  (\(dt)  dt[, .(number_of_trips = comma(number_of_trips),
                 trips_with_tips = comma(trips_with_tips),
                 trips_with_tips_pct = percent(trips_with_tips / number_of_trips),
                 driver_net_earning = dollar(driver_net_earning),
                 tips = dollar(tips),
                 tips_pct = percent(tips/driver_net_earning))
          ][, melt(.SD, 
                   measure.vars = names(.SD),
                   variable.name = "Summary Variable",
                   value.name = "Total",
                   variable.factor = FALSE)]
   )()

```

## Project's Side Benefits

It's also important to consider that Taxi companies and customers can both benefit from drivers earning more tips in several ways:

1. **Taxi Companies**: 
    - **Employee Satisfaction**: Higher tips can lead to increased job satisfaction among drivers, which can improve their performance and reduce turnover rates.
    - **Company Reputation**: If drivers are earning more tips, it could indicate that they are providing excellent service, which can enhance the company's reputation.
    - **Customer Retention**: Satisfied drivers are more likely to provide better customer service, which can lead to higher customer retention rates.

2. **Customers**:
    - **Better Service**: Drivers who earn more tips are often those who provide better service. This could mean cleaner vehicles, more courteous behavior, and a more enjoyable ride overall.
    - **Driver Availability**: If the tip earnings are high, it could attract more drivers to work, potentially reducing wait times for customers.
    - **Safety**: Drivers who are not worried about their earnings might be less likely to engage in risky behaviors (like speeding or working overly long shifts) to earn more.
    
    
## Defining Outcome to Predict

Once we have a clear business problem to solve, it's important to also make the translation into a *data science* problem that we can solve using the available data.

As we want to increase earning and reduce time, we can use calculate and predict the `profit_rate` of each trip, to understand **what is behind the best pay trips**.

$$
\text{profit rate} = \frac{\text{driver pay} + \text{tips}}{\text{dropoff datetime}-\text{request datetime}}
$$



## Deliverables

A **Shiny app** which assists the drivers in focusing their attention on the better trips.


```
-- Crear una tabla temporal final para almacenar los resultados
CREATE TEMP TABLE better_trip_validation AS
WITH trip_performance AS (
    SELECT 
        ROW_NUMBER() OVER () AS trip_id,
        *,
        (driver_pay + tips) / (EXTRACT(EPOCH FROM (dropoff_datetime - request_datetime)) / 3600) AS performance_per_hour
    FROM 
        trips
),
better_trips AS (
    SELECT 
        t1.trip_id AS trip_id,
        CASE WHEN EXISTS (
            SELECT 1
            FROM trip_performance t2
            WHERE t2.trip_id <> t1.trip_id
              AND t2.request_datetime > t1.request_datetime
              AND t2.request_datetime <= t1.request_datetime + INTERVAL '20 minutes'
              AND (t2.driver_pay + t2.tips) / (EXTRACT(EPOCH FROM (t2.dropoff_datetime - t1.request_datetime)) / 3600) > t1.performance_per_hour
        ) THEN 1 ELSE 0 END AS better_trip_20_m
    FROM 
        trip_performance t1
)

-- Unir los resultados con los detalles de cada viaje
SELECT *
FROM trip_performance t1
JOIN better_trips t2
ON t1.trip_id = t2.trip_id;
```
